---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setup}
library(knitr)
knitr::opts_knit$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "E:\\Coursera\\Reproducible_Research\\RepData_PeerAssessment1-master\\Data\\")
```


## Loading and preprocessing the data
```{r}
library(data.table)
amdata <- read.table("activity.csv", header = TRUE,sep=",",stringsAsFactors = FALSE)
head(amdata)
```

### Structure of Activity Management Data
```{r}
str(amdata)
```

### Convert date field to Date format
```{r}
amdata$date <- as.Date(amdata$date)
str(amdata)
```

### Summary of Activity Data
```{r}
summary(amdata)
```

## What is mean total number of steps taken per day?

## Daily total steps:
```{r}
library(dplyr)
library(ggplot2)
library(xtable)

daily_steps<- group_by(amdata, date)%>% summarise_each(funs(sum(steps, na.rm=TRUE))) %>% select("Date" = date, "Daily_total_steps"= steps)

kable(head(daily_steps))
```

### Histogram of the total number of steps taken each day
```{r}
ggplot(amdata, aes(date, steps ))+geom_bar(stat="identity", fill="steelblue")+
  theme(axis.text.x = element_text(angle =110, vjust = .5, color = "gray")) + 
  xlab("Date") + 
  ylab("Steps") + 
  ggtitle("Total number of steps taken each day")
```

### Average daily steps:
```{r }
daily_mean <- round(mean(daily_steps$Daily_total_steps, na.rm=TRUE))
```
Average daily steps: `r daily_mean`

### Daily median steps:
```{r}
library(xtable)
daily_median_steps <- median(daily_steps$Daily_total_steps, na.rm=TRUE)
```
Daily median steps: `r daily_median_steps`

## What is the average daily activity pattern?
```{r}
library(dplyr)
library(ggplot2)
interval_steps<- group_by(amdata, interval)%>% summarise_each(funs(mean(steps, na.rm=TRUE))) %>% select("Interval"=interval, "Average_steps"= steps)
max_interval_steps <- max(interval_steps$Average_steps, na.rm=TRUE)
max_interval <- interval_steps[which(interval_steps$Average_steps == max_interval_steps),]$Interval
ggplot(interval_steps, aes(x = Interval, y = Average_steps)) +
         geom_line()+
  xlab("Interval") + 
  ylab("Number of steps") + 
  ggtitle("Average daily activity pattern")
```

5-minute Interval `r max_interval` has maximum number of steps: `r max_interval_steps`

## Imputing missing values
### Missing values will be imputed using mice package
1. m  - Refers to number of imputed data sets  
2. maxit - Refers to no. of iterations taken to impute missing values  
3. method - Refers to method used in imputation  
4. we will use (pmm) predictive mean matching as steps is a numeric variable  
5. We will adda a new variable (isteps- imputed steps) to amdata
```{r}
library(mice)
amdata_miss <- select(amdata, c(1,3))  # only select numerical variables
am1 <- mice(amdata_miss, m=1, maxit = 5, method = 'pmm', seed = 500)
amdata_miss <- complete(am1,1)
amdata$isteps <- amdata_miss$steps
```

### Histogram of the total number of steps taken each day (with imputed vales for steps)
```{r}
ggplot(amdata, aes(date, isteps ))+geom_bar(stat="identity", fill="steelblue")+
  theme(axis.text.x = element_text(angle =110, vjust = .5, color = "gray")) + 
  xlab("Date") + 
  ylab("Steps") + 
  ggtitle("Total number of steps taken each day")
```

### Average daily steps(with imputed values):
```{r }
daily_isteps<- group_by(amdata, date)%>% summarise_each(funs(sum(isteps, na.rm=TRUE))) %>% select("Date" = date, "Daily_total_steps"= isteps)
daily_imean <- round(mean(daily_steps$Daily_total_steps, na.rm=TRUE))
```
Average daily steps with imputation: `r daily_imean`  
Average daily steps without imputation: `r daily_mean`  

### Daily median steps:
```{r}
daily_median_isteps <- median(daily_steps$Daily_total_steps, na.rm=TRUE)
```
Daily median steps with imputation: `r daily_median_isteps`  
Daily median steps without imputation: `r daily_median_steps` 

## Are there differences in activity patterns between weekdays and weekends?

```{r}
# Add one factor variable "daytype" to amdata
amdata$daytype <- ifelse(weekdays(amdata$date)=="Sunday"|weekdays(amdata$date) == "Saturday","Weekend","Weekday")
amdata$daytype <- as.factor(amdata$daytype)
interval_isteps<- group_by(amdata, interval,daytype)%>% summarise_each(funs(mean(isteps, na.rm=TRUE)))
```

```{r}
# Use ggplot2 to display a panal plot of time series
ggplot(interval_isteps, aes(x = interval, y = isteps, col = daytype)) +
  geom_line() + facet_grid( ~ daytype) +
  xlab("Interval") + 
  ylab("Number of steps") + 
  ggtitle("Activity pattern between weekdays and weekends")
```

